cardinalR: Generating Interesting High-Dimensional Data Structures

### 2025-12-18 paper-cardinalR.Rmd
Updated Table 2

@@

                              "Möbius strip in $3\text{-}D$.",
                              "Quadratic pattern in $2\text{-}D$.",
                                "Cubic pattern in $2\text{-}D$.",

@@

```{r}
arg_branching_tb <- tibble(arg = c("n", "k", "p", "h", "ratio", "s", "r", "n_vec", "k_small", "r_vec", "spe", "w", "steps", "spins", "hc", "non_fac"), 
                          type = c("integer", "integer", "integer", "real value",
                                   "real value", "real value", "real value", "integers", "integer", "real values", "real value", "real value", "integer", "integer", "real value", "real value"),
                        exp = c("Number of points.",
                                "Number of clusters.",
                                "Number of dimensions.",
                                "Height.",
                                "Radius tip to radius base ratio.",
                                "Variance-covariance matrix.",
                                "Radius.",
                                "Sample sizes of the big and small spheres",
                                "Number of small spheres.",
                                "Radius of the big and small spheres.",
                                "How far apart the small spheres are placed.",
                                "Vertical variation",
                                "Number of steps for the theta parameter.",
                                "Number of loops of the spiral.",
                                "Steepness and vertical scaling of the hyperbola.",
                                "Strength of this sinusoidal effect.")) |> 
   mutate(arg = gsub("_", "\\\\_", arg))
```

@@

The remaining dimensions are based on the specific pyramid shape. For the rectangular based pyramid, `gen_pyrrect(n, p, h, l_vec, rt)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:pyr) a"} else { "Figure \\@ref(fig:pyr-proj) a"})`) the base shape is a rectangle whose size shrinks linearly with height. Let $l_x$ and $l_y$ denote the half-widths of the rectangular base in the $X_1$ and $X_2$ directions, specified via $l=(l_x,l_y)$, and let $r_t$ denote the half-width at the pyramid tip. At height $z\in [0,h]$, the half-widths of the rectangular cross-section are $r_x(z) = r_t + (l_x - r_t)z/h$, $r_y(z) = r_t + (l_y - r_t)z/h$. The first three coordinates are then defined as $X_1 \sim U(-r_x(z),\ r_x(z)), \quad X_2 \sim U(-r_y(z),\ r_y(z)),\text{ and }X_3 \sim U(-r_x(z),\ r_x(z))$.

>>>>

                             "n, p, l_vec, rt",
                             "n, p, h, l, rt",
                             "n, p, h, rb",

@@

                              "Möbius strip in $3\\text{-}D$.",
                              "Quadratic pattern in $2\\text{-}D$.",
                                "Cubic pattern in $2\\text{-}D$.",

@@

```{r}
arg_branching_tb <- tibble(arg = c("n", "k", "p", "h", "ratio", "s", "r", "n_vec", "k_small", "r_vec", "spe", "w", "steps", "spins", "hc", "non_fac", "l", "l_vec", "rt", "rb"), 
                          type = c("integer", "integer", "integer", "real value",
                                   "real value", "real value", "real value", "integers", "integer", "real values", "real value", "real value", "integer", "integer", "real value", "real value", "real value", "real values", "real value", "real value"),
                        exp = c("Number of points.",
                                "Number of clusters.",
                                "Number of dimensions.",
                                "Height.",
                                "Radius tip to radius base ratio.",
                                "Variance-covariance matrix.",
                                "Radius.",
                                "Sample sizes of the big and small spheres",
                                "Number of small spheres.",
                                "Radius of the big and small spheres.",
                                "How far apart the small spheres are placed.",
                                "Vertical variation",
                                "Number of steps for the theta parameter.",
                                "Number of loops of the spiral.",
                                "Steepness and vertical scaling of the hyperbola.",
                                "Strength of this sinusoidal effect.",
                                "Base length of the pyramid.",
                                "Base lengths along the and y of the pyramid.",
                                "Tip radius of the pyramid.",
                                "Base radius of the pyramid.")) |> 
   mutate(arg = gsub("_", "\\\\_", arg))
```

@@

The remaining dimensions are based on the specific pyramid shape. For the rectangular based pyramid, `gen_pyrrect(n, p, h, l_vec, rt)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:pyr) a"} else { "Figure \\@ref(fig:pyr-proj) a"})`) the base shape is a rectangle whose size shrinks linearly with height. Let $l_x$ and $l_y$ denote the half-widths of the rectangular base in the $X_1$ and $X_2$ directions, specified via $l_{vec}=(l_x,l_y)$, and let $r_t$ denote the half-width at the pyramid tip. At height $z\in [0,h]$, the half-widths of the rectangular cross-section are $r_x(z) = r_t + (l_x - r_t)z/h$, $r_y(z) = r_t + (l_y - r_t)z/h$. The first three coordinates are then defined as $X_1 \sim U(-r_x(z),\ r_x(z)), \quad X_2 \sim U(-r_y(z),\ r_y(z)),\text{ and }X_3 \sim U(-r_x(z),\ r_x(z))$.

### 2025-12-18 paper-cardinalR.tex
Updated Table 2

\caption{\label{tab:shape-tb-pdf}Overview of shape-generation functions, including their required parameters and a brief description of each geometric structure produced. The generators cover branching patterns, spheres, spirals, pyramids, Gaussian clouds, and other nonlinear shapes.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3.5cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Arguments & Explanation\\
\midrule
\texttt{gen\_expbranches} & \texttt{n, k} & Exponential shaped branches.\\
\texttt{gen\_linearbranches} & \texttt{n, k} & Linear shaped branches.\\
\texttt{gen\_curvybranches} & \texttt{n, k} & Curvy shaped branches.\\
\texttt{gen\_orglinearbranches} & \texttt{n, p, k} & Linear shaped branches originated in one point.\\
\texttt{gen\_orgcurvybranches} & \texttt{n, p, k} & Curvy shaped branches originated in one point.\\
\texttt{gen\_cone} & \texttt{n, p, h, ratio} & Cone-shaped structure.\\
\texttt{gen\_gridcube} & \texttt{n, p} & Cube with specified grid points along each axes.\\
\texttt{gen\_unifcube} & \texttt{n, p} & Cube with uniform points.\\
\texttt{gen\_gaussian} & \texttt{n, p, s} & Multivariate Gaussian cloud.\\
\texttt{gen\_longlinear} & \texttt{n, p} & Long linear structure.\\
\texttt{gen\_mobius} & \texttt{n} & Möbius strip in $3  ext{-}D$.\\
\texttt{gen\_quadratic} & \texttt{n} & Quadratic pattern in $2  ext{-}D$.\\
\texttt{gen\_cubic} & \texttt{n} & Cubic pattern in $2  ext{-}D$.\\
\texttt{gen\_pyrrect} & \texttt{n, p} & Rectangular-base, with a sharp or blunted apex.\\
\texttt{gen\_pyrtri} & \texttt{n, p} & Triangular-base, with a sharp or blunted apex.\\
\texttt{gen\_pyrstar} & \texttt{n, p} & Star-shaped base, with a sharp or blunted apex.\\
\texttt{gen\_pyrfrac} & \texttt{n, p} & Pyramid with triangular pyramid-shaped holes.\\
\texttt{gen\_scurve} & \texttt{n} & S-curve in $3   ext{-}D$.\\
\texttt{gen\_circle} & \texttt{n, p} & Circle.\\
\texttt{gen\_curvycycle} & \texttt{n, p} & Curvy cell cycle.\\
\texttt{gen\_unifsphere} & \texttt{n, r} & Uniform ball.\\
\texttt{gen\_hollowsphere} & \texttt{n, p} & Hollow sphere.\\
\texttt{gen\_gridedsphere} & \texttt{n} & Grided sphere.\\
\texttt{gen\_clusteredspheres} & \texttt{n, k, r, loc} & Multiple small spheres within a big sphere.\\
\texttt{gen\_hemisphere} & \texttt{n, p} & Hemisphere.\\
\texttt{gen\_swissroll} & \texttt{n, w} & Swissroll structure.\\
\texttt{gen\_trefoil4d} & \texttt{n, steps} & Trefoil in $4\text{-}D$.\\
\texttt{gen\_trefoil3d} & \texttt{n, steps} & Trefoil in $3\text{-}D$.\\
\texttt{gen\_crescent} & \texttt{n} & Crescent pattern.\\
\texttt{gen\_curvycylinder} & \texttt{n, h} & Curvy cylinder.\\
\texttt{gen\_sphericalspiral} & \texttt{n, spins} & Spherical spiral.\\
\texttt{gen\_helicalspiral} & \texttt{n} & Helical spiral.\\
\texttt{gen\_conicspiral} & \texttt{n, spins} & Conic spiral.\\
\texttt{gen\_nonlinear} & \texttt{n, hc, non\_fac} & Nonlinear hyperbola.\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:arg-shape-tb-pdf}Argument definitions for the shape generators. The table lists each argument, its data type, and a description of its role in controlling geometric structure, dimensionality, scaling, curvature, spacing, and other features of the simulated high-dimensional datasets.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{3cm}l}
\toprule
Argument & Type (positive) & Explanation\\
\midrule
\texttt{n} & integer & Number of points.\\
\texttt{k} & integer & Number of clusters.\\
\texttt{p} & integer & Number of dimensions.\\
\texttt{h} & real value & Height.\\
\texttt{ratio} & real value & Radius tip to radius base ratio.\\
\addlinespace
\texttt{s} & real value & Variance-covariance matrix.\\
\texttt{r} & real value & Radius.\\
\texttt{n\_vec} & integers & Sample sizes of the big and small spheres\\
\texttt{k\_small} & integer & Number of small spheres.\\
\texttt{r\_vec} & real values & Radius of the big and small spheres.\\
\addlinespace
\texttt{spe} & real value & How far apart the small spheres are placed.\\
\texttt{w} & real value & Vertical variation\\
\texttt{steps} & integer & Number of steps for the theta parameter.\\
\texttt{spins} & integer & Number of loops of the spiral.\\
\texttt{hc} & real value & Steepness and vertical scaling of the hyperbola.\\
\addlinespace
\texttt{non\_fac} & real value & Strength of this sinusoidal effect.\\
\bottomrule
\end{tabular}
\end{table}

@@

The remaining dimensions are based on the specific pyramid shape. For the rectangular based pyramid, \texttt{gen\_pyrrect(n,\ p,\ h,\ l\_vec,\ rt)} (Figure \ref{fig:pyr-proj} a) the base shape is a rectangle whose size shrinks linearly with height. Let \(l_x\) and \(l_y\) denote the half-widths of the rectangular base in the \(X_1\) and \(X_2\) directions, specified via \(l=(l_x,l_y)\), and let \(r_t\) denote the half-width at the pyramid tip. At height \(z\in [0,h]\), the half-widths of the rectangular cross-section are \(r_x(z) = r_t + (l_x - r_t)z/h\), \(r_y(z) = r_t + (l_y - r_t)z/h\). The first three coordinates are then defined as \(X_1 \sim U(-r_x(z),\ r_x(z)), \quad X_2 \sim U(-r_y(z),\ r_y(z)),\text{ and }X_3 \sim U(-r_x(z),\ r_x(z))\).

>>>>

\caption{\label{tab:shape-tb-pdf}Overview of shape-generation functions, including their required parameters and a brief description of each geometric structure produced. The generators cover branching patterns, spheres, spirals, pyramids, Gaussian clouds, and other nonlinear shapes.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3.5cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Arguments & Explanation\\
\midrule
\texttt{gen\_expbranches} & \texttt{n, k} & Exponential shaped branches.\\
\texttt{gen\_linearbranches} & \texttt{n, k} & Linear shaped branches.\\
\texttt{gen\_curvybranches} & \texttt{n, k} & Curvy shaped branches.\\
\texttt{gen\_orglinearbranches} & \texttt{n, p, k} & Linear shaped branches originated in one point.\\
\texttt{gen\_orgcurvybranches} & \texttt{n, p, k} & Curvy shaped branches originated in one point.\\
\texttt{gen\_cone} & \texttt{n, p, h, ratio} & Cone-shaped structure.\\
\texttt{gen\_gridcube} & \texttt{n, p} & Cube with specified grid points along each axes.\\
\texttt{gen\_unifcube} & \texttt{n, p} & Cube with uniform points.\\
\texttt{gen\_gaussian} & \texttt{n, p, s} & Multivariate Gaussian cloud.\\
\texttt{gen\_longlinear} & \texttt{n, p} & Long linear structure.\\
\texttt{gen\_mobius} & \texttt{n} & Möbius strip in $3\text{-}D$.\\
\texttt{gen\_quadratic} & \texttt{n} & Quadratic pattern in $2\text{-}D$.\\
\texttt{gen\_cubic} & \texttt{n} & Cubic pattern in $2\text{-}D$.\\
\texttt{gen\_pyrrect} & \texttt{n, p, l\_vec, rt} & Rectangular-base, with a sharp or blunted apex.\\
\texttt{gen\_pyrtri} & \texttt{n, p, h, l, rt} & Triangular-base, with a sharp or blunted apex.\\
\texttt{gen\_pyrstar} & \texttt{n, p, h, rb} & Star-shaped base, with a sharp or blunted apex.\\
\texttt{gen\_pyrfrac} & \texttt{n, p} & Pyramid with triangular pyramid-shaped holes.\\
\texttt{gen\_scurve} & \texttt{n} & S-curve in $3   ext{-}D$.\\
\texttt{gen\_circle} & \texttt{n, p} & Circle.\\
\texttt{gen\_curvycycle} & \texttt{n, p} & Curvy cell cycle.\\
\texttt{gen\_unifsphere} & \texttt{n, r} & Uniform ball.\\
\texttt{gen\_hollowsphere} & \texttt{n, p} & Hollow sphere.\\
\texttt{gen\_gridedsphere} & \texttt{n} & Grided sphere.\\
\texttt{gen\_clusteredspheres} & \texttt{n, k, r, loc} & Multiple small spheres within a big sphere.\\
\texttt{gen\_hemisphere} & \texttt{n, p} & Hemisphere.\\
\texttt{gen\_swissroll} & \texttt{n, w} & Swissroll structure.\\
\texttt{gen\_trefoil4d} & \texttt{n, steps} & Trefoil in $4\text{-}D$.\\
\texttt{gen\_trefoil3d} & \texttt{n, steps} & Trefoil in $3\text{-}D$.\\
\texttt{gen\_crescent} & \texttt{n} & Crescent pattern.\\
\texttt{gen\_curvycylinder} & \texttt{n, h} & Curvy cylinder.\\
\texttt{gen\_sphericalspiral} & \texttt{n, spins} & Spherical spiral.\\
\texttt{gen\_helicalspiral} & \texttt{n} & Helical spiral.\\
\texttt{gen\_conicspiral} & \texttt{n, spins} & Conic spiral.\\
\texttt{gen\_nonlinear} & \texttt{n, hc, non\_fac} & Nonlinear hyperbola.\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:arg-shape-tb-pdf}Argument definitions for the shape generators. The table lists each argument, its data type, and a description of its role in controlling geometric structure, dimensionality, scaling, curvature, spacing, and other features of the simulated high-dimensional datasets.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{3cm}l}
\toprule
Argument & Type (positive) & Explanation\\
\midrule
\texttt{n} & integer & Number of points.\\
\texttt{k} & integer & Number of clusters.\\
\texttt{p} & integer & Number of dimensions.\\
\texttt{h} & real value & Height.\\
\texttt{ratio} & real value & Radius tip to radius base ratio.\\
\addlinespace
\texttt{s} & real value & Variance-covariance matrix.\\
\texttt{r} & real value & Radius.\\
\texttt{n\_vec} & integers & Sample sizes of the big and small spheres\\
\texttt{k\_small} & integer & Number of small spheres.\\
\texttt{r\_vec} & real values & Radius of the big and small spheres.\\
\addlinespace
\texttt{spe} & real value & How far apart the small spheres are placed.\\
\texttt{w} & real value & Vertical variation\\
\texttt{steps} & integer & Number of steps for the theta parameter.\\
\texttt{spins} & integer & Number of loops of the spiral.\\
\texttt{hc} & real value & Steepness and vertical scaling of the hyperbola.\\
\addlinespace
\texttt{non\_fac} & real value & Strength of this sinusoidal effect.\\
\texttt{l} & real value & Base length of the pyramid.\\
\texttt{l\_vec} & real values & Base lengths along the and y of the pyramid.\\
\texttt{rt} & real value & Tip radius of the pyramid.\\
\texttt{rb} & real value & Base radius of the pyramid.\\
\bottomrule
\end{tabular}
\end{table}

@@

The remaining dimensions are based on the specific pyramid shape. For the rectangular based pyramid, \texttt{gen\_pyrrect(n,\ p,\ h,\ l\_vec,\ rt)} (Figure \ref{fig:pyr-proj} a) the base shape is a rectangle whose size shrinks linearly with height. Let \(l_x\) and \(l_y\) denote the half-widths of the rectangular base in the \(X_1\) and \(X_2\) directions, specified via \(l_{vec}=(l_x,l_y)\), and let \(r_t\) denote the half-width at the pyramid tip. At height \(z\in [0,h]\), the half-widths of the rectangular cross-section are \(r_x(z) = r_t + (l_x - r_t)z/h\), \(r_y(z) = r_t + (l_y - r_t)z/h\). The first three coordinates are then defined as \(X_1 \sim U(-r_x(z),\ r_x(z)), \quad X_2 \sim U(-r_y(z),\ r_y(z)),\text{ and }X_3 \sim U(-r_x(z),\ r_x(z))\).
